---
title: "RNA-seq Data"
format: 
  html:
    theme: 
      light: flatly
      dark: darkly
execute:
  echo: false
editor: visual
---

```{r}
#| inclue: false
#| message: false
library(readr)
library(dplyr)
library(tidyr)

library(ggplot2)
library(cowplot)
library(hexbin)
theme_set(theme_cowplot())

source('https://tinyurl.com/wjm6ryj')

if(FALSE)
{
    download.file(url = "https://github.com/Bioconductor/bioconductor-teaching/raw/master/data/GSE96870/rnaseq.csv",
              destfile = "data/rnaseq.csv")
}

### Read in data here ###
dat <- read_csv('data/rnaseq.csv') %>%
    mutate(l2exp = log2(expression + 1))
```

## Agenda

-   Today's lab
    -   RNA-Seq data visualization - this lab borrows heavily from a [Carpentries incubator module on visualization](https://carpentries-incubator.github.io/bioc-intro/40-visualization.html).

## RNA-Seq data

To start out with, fork the [starter code](https://classroom.github.com/a/Yz2GMIqU) and clone the repository to your local machine. Once we have done that, let's explore the data provided:

```{r}
dat
```

Take a look at the distribution of the expression levels:

```{r}
#| message: false
(g1 <- )
## You'll probably notice an automatic message that appears when drawing the histogram: `stat_bin()` using `bins = 30`. Pick better value with `binwidth`.
## Change the arguments bins or binwidth of geom_histogram() to change the number or width of the bins.
```

The data should be transformed. Transforming on the $\log_{10}$ scale is easy in `ggplot2`,

```{r}
#| warning: false
g1 + 
```

but we often use a $log_2$ scale when looking at expression data:

```{r}
#| warning: false
ticks_at <- c(0,5,10,15)
ggplot(dat, aes(l2exp)) +

```

### Building plots iteratively

We will now draw a scatter plot with two continuous variables and the `geom_point()` function. This graph will represent the $\log_2$ fold changes of expression comparing time 8 versus time 0, and time 4 versus time 0. To this end, we first need to compute the means of the log-transformed expression values by gene and time, then the log fold changes by subtracting the mean log expressions between time 8 and time 0 and between time 4 and time 0.

```{r}
#| message: false
# fold change data
dat_fc <- dat %>% 
    select(gene, 
           time,
           gene_biotype,
           expression) %>%
    
     # calculate summaries of log2 expression grouped by gene, time, gene_biotype
    group_by(gene, time, gene_biotype) %>%
    summarize(mean_exp = mean(log2(expression + 1))) %>%
    ungroup() %>%
    
    # pivot wider so that there is a separate column for each value in time
    pivot_wider(names_from = time,
                values_from = mean_exp) %>%
    
    mutate(time_8_vs_0 = `8` - `0`, # compare time 8 vs 0
           time_4_vs_0 = `4` - `0`) # compare time 4 vs 0

# visualize time_4_vs_0 by time_8_vs_0 
ggplot(dat_fc, ...) +

```

We could also add color, but it begins to get a little messy with the large number of categories we have.

```{r}
ggplot(dat_fc, aes(time_4_vs_0, time_8_vs_0)) +

```

Over-plotting can be an issue with large datasets like this. One solution is using `hexbin:geom_hex()`.

```{r}
ggplot(dat_fc, aes(time_4_vs_0, time_8_vs_0)) +

```

### Box plots

We can use boxplots to visualize the distribution of gene expressions within each sample:

```{r}
ggplot(dat, ) +

```

Adding points to box plots can give us a better understanding of the underlying distributions.

```{r}
ggplot(dat, aes(sample, l2exp)) +
    geom_jitter(alpha = 0.02, width = .15, color = cbbPalette[2]) +

```

We could also use `geom_violin` to better see the distribution of points. Try coloring the points by `time` as well.

```{r}
ggplot(dat, aes(sample, l2exp)) +
    geom_jitter(alpha = 0.02, width = .15, aes(color = as.factor(time))) +

```

### Line plots

Let's calculate the mean expression per duration of the infection for the 10 genes having the highest log fold changes comparing time 8 versus time 0. We can do this either by coloring each gene or using `facet_wrap`.

```{r}
#| message: false

# First, we need to select the genes
genes_selected <- arrange(dat_fc, desc(time_8_vs_0))$gene[1:10]

# Then create a subset of dat containing the 10 selected genes
mean_exp_by_time <- dat %>%
    filter(gene %in% genes_selected) %>%
    
    # Then group the data by gene and time
    group_by(gene, time) %>%
    # calculate the mean gene expression within each group
    summarize(mean_exp = mean(l2exp)) %>%
    ungroup()

ggplot(mean_exp_by_time, ) +

ggplot(mean_exp_by_time, ) +

```

Now we would like to split the line in each plot by the sex of the mice.

```{r}
#| message: false
# Then create a subset of dat containing the 10 selected genes
mean_exp_by_time_sex <- dat %>%
    filter(gene %in% genes_selected) %>%
    
    # Then group the data by gene, time, and sex
    group_by(gene, time, sex) %>%
    # calculate the mean gene expression within each group
    summarize(mean_exp = mean(l2exp)) %>%
    ungroup()

ggplot(mean_exp_by_time_sex, ) +

```

Let's do something similar and create a plot that depicts how the average expression of each chromosome changes through the duration of infection.

```{r}
#| message: false
mean_exp_by_chromosome <- dat %>%
    group_by(chromosome_name, time) %>%
    summarize(mean_exp = mean(l2exp))

ggplot(mean_exp_by_chromosome, ) +

```

The `facet_wrap` geometry extracts plots into an arbitrary number of dimensions to allow them to cleanly fit on one page. On the other hand, the `facet_grid` geometry allows you to explicitly specify how you want your plots to be arranged via formula notation (`rows ~ columns`; a `.` can be used as a placeholder that indicates only one row or column).

Let's modify the previous plot to compare how the mean gene expression of males and females has changed through time:

```{r}
# One column, facet by rows
ggplot(mean_exp_by_time_sex, ) +

```
